---
title: "Homework 3 Quarto"
format: pdf
editor: visual
---

<https://github.com/mateomoribaez-design/ENVS-193DS_homework-03.git>

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(here)
library(janitor)
library(readxl)  # lead in packages in setup chunk
```

```{r}
# store salinity data
salinity <- read_csv(
  here("data", "salinity-pickleweed.csv"))

# store personal data
my_data <- read_csv(
  here("data", "my_data.csv"))
```

### Problem 1

#### a.

The two appropriate tests to use in this scenario are the Pearson correlation and Simple linear regression tests. The pearson correlation test measures strength of association while the simple linear regression test models the response variable as a function of the predictor variable (biomass and salinity respectively).

#### b.

```{r}

ggplot(salinity,
      aes(x = salinity_mS_cm,
        y = pickleweed)) +
  geom_point(color = "darkgreen", size = 3) +
  geom_smooth(method = "lm",
              se = TRUE,
              color = "steelblue") +
  labs(
    x = "Soil Salinity (mS/cm)",
    y = "Pickleweed Biomass (g)",
    title = "Relationship Between Soil Salinity and Pickleweed Biomass"
  ) +
  theme_minimal()


```

#### c and d

#### Checking assumptions

Linearity

```{r}
plot(salinity$salinity_mS_cm, # create scatterplot of soil salinity vs pickleweed biomass
     salinity$pickleweed,
     xlab = "Soil Salinity (mS/cm)", # label x and y axis
     ylab = "Biomass (g)")
model <- lm(pickleweed ~ salinity_mS_cm, data = salinity) # fit linear regression model predicting biomass from salinity
abline(model, col = "red") # add fitted regression line to scatterplot
```

I checked the linearity assumption by plotting soil salinity against pickleweed biomass with a scatterplot and a fitted linear regression line. The scatterplot shows an increasing trend with salinity, which suggests that the relationship between salinity and biomass is somewhat linear. I think the linearity assumption is satisfied.

Normality

```{r}
qqnorm(residuals(model)) # create qq plot of the model residuals to visually check if they follow a normal distribution
qqline(residuals(model), col = "blue") # Add reference line to qqplot

shapiro.test(residuals(model)) # use a shapiro-wilk test for normality
```

I checked for normality using both a qqplot and a shapiro-wilk test. The qq plot shows that most data points fall close to the reference line, indicating that the residuals are decently normally distributed. I think that together, these results show that the normality assumption is met.

Homoscedasticity

```{r}
plot(fitted(model), residuals(model), # plot fitted values against model residuals
     xlab = "Fitted Values", # predicted biomass values from the model
     ylab = "Residuals") # difference between observed and predicted values
abline(h = 0, col = "red") # add horizontal reference line
```

I checked for homoscedasticity by plotting residuals against fitted values from the regression model. The points are randomly scattered around the horizontal reference line, without a clear pattern, which suggests that the variance of residuals is constant across predicted values. I think the homoscedasticity assumption is met.

#### Run the test

Pearson correlation

```{r}
cor.test(salinity$salinity_mS_cm, # use a Pearson correlation test to measure strength and direction of the relationship between salinity and biomass
         salinity$pickleweed,
         method = "pearson")
```

To evaluate the strength of the relationship between soil salinity and pickleweed biomass, I used a Pearson correlation test because both variables are continuous. The test output showed a moderate positive correlation between salinity and biomass (r = 0.534, t = 2.898, df = 21, p = 0.0086). The 95% confidence interval for the correlation coefficient ranged from 0.157 to 0.758. With a p-value less than 0.05, we reject the null hypothesis and conclude that there is a significant positive relationship between soil salinity and pickleweed biomass.

Linear regression (same from above)

```{r}
summary(model) # display a summary of the linear regression model
```

I use simple linear regression to model pickleweed biomass as a function of soil salinity. This test shows us how much biomass changes with increasing salinity. The regression outputs show that soil salinity significantly predicts pickleweed biomass (β = 2.083, SE = 0.719, t = 2.898, p = 0.0086). The model intercept was 13.09, showing the predicted biomass when the salinity is zero, and there was a 28.6% variation in biomass (R² = 0.286). Overall, the model was statistically significant (F(1,21) = 8.398, p = 0.0086).

#### e.

Our test outcomes suggest that soil salinity significantly affects pickleweed biomass at our restoration site. If biomass declines with increasing salinity, this means that highly saline areas could reduce plant growth. For restoration planning, this means we should plant in moderately saline zones, make soil adjustments for high salinity areas, and monitor the salinity levels in these areas before planting to ensure success.

#### f. 

I might have ran both tests when I shouldn't have but I think I'll leave it in and just answer the second part of 1f.

Both Pearson correlation and linear regression were conducted, and they both evaluate the same null hypothesis:\
H₀: There is no linear relationship between salinity and biomass

Because linear regression testing and pearson correlation are equivalent in simpe linear regression, both tests ended up leading to the same comclusion about statistical significance. From the pearson correlation test, the p-value was less than 0.05, which means that we reject the null hypothesis and conclude that there is a significant positive relationship between soil salinity and pickleweed biomass. The linear regression analysis showed that soil salinity significantly predicts pickleweed biomass, which follows what the Pearson's correlation outputs showed. Above in the descriptions, outputs and interpretation of the tests I have my outputs specified.

### Problem 2

#### a. 

Categorical predictor visualization

```{r}
my_data <- read_csv(here("data", "my_data.csv")) |> 
  clean_names() |> # load and clean my data
  rename(ocean_activity = contains("ocean_activity")) |> # give my long ocean activity column a short and usable name
  filter(!ocean_activity2 %in% c("none")) |> # remove rows where the activity is none because none is a significant part of the data
  mutate(
    daily_workload_hours = as.numeric(daily_workload_hours), # workload column is treated as a number for the math in this chunk
    workload_cat = case_when( # create new column based on the hour ranges listed below
      daily_workload_hours <= 3 ~ "Low (< 3 hours)",
      daily_workload_hours <= 7 ~ "Medium (3-7 hours)",
      daily_workload_hours > 7 ~ "High (>7 hours)"
    ),
    workload_cat = factor(workload_cat, levels = c("Low (< 3 hours)", "Medium (3-7 hours)", "High (>7 hours)")) # convert workload_cat to a factor
  )

ggplot(my_data, aes(x = workload_cat, fill = ocean_activity2)) + # create histogram visualization
  geom_bar(position = "dodge") + # create side-by-side bars for each activity within the workload groups
  labs( # add descriptive labels to the title and axes
    title = "Ocean Activities by Daily Workload",
    x = "Daily Workload Category (hours)",
    y = "Count of Ocean Activities",
    fill = "Ocean Activity"
  ) +
  scale_fill_brewer(palette = "Set2") + # apply palette
  theme_minimal() # change background theme
```

Continuous predictor visualization

```{r}
my_jitter_data <- read_csv(here("data", "my_data.csv")) |> # read in data
  clean_names() |> # clean column names
  rename(ocean = contains("ocean_activity")) |> # Rename the long activity column to 'ocean' for the y-axis
  mutate(
    ocean = case_when(
      ocean %in% c("none", "None", "No", "no") ~ "No", # if the value is "none", "None", or "No", label it as "No"
      TRUE ~ "Yes" # anything other than no shall be yes
    ),
    daily_workload_hours = as.numeric(daily_workload_hours)) # ensure workload column is treated as numerical values
ggplot(data = my_jitter_data, # create scatterplot visualization
       aes(x = daily_workload_hours, y = ocean)) + # use the cleaned up column names
  geom_jitter( # add jittered points to scatterplot
    width = 0,   # no horizontal jitter, we want the points to stay on their daily workload hours
    height = 0.1 # jitter points slightly in the vertical direction
  ) +
  theme_minimal() + # change theme
  labs(
    title = "Ocean-Going against Daily Workload", # use adequate x and y axes labels
    x = "Daily Workload (hours)",
    y = "Went in Ocean (Yes/No)"
  )
```

#### b.

Figure 1: Categorical Predictor Visualization

Figure 1: Frequency of Ocean Activity Types by Daily Workload Categories. This bar chart shows how different types of ocean activity are distributed across three different workload levels: Low( \< 3 hrs), Medium (3-7 hrs), and High ( \> 7 hrs). The chart shows that the highest volume and variety of ocean activities occurs on low workload days, with longboarding being the most frequent activity in that category. On the other hand, as workload hours increase, ocean activity is only quick dips.

Figure 2: Continuous Predictor Visualization

Figure 2: Daily Ocean Participation Relative to Continuous Workload Hours. This scatter plot shows the relationship between the specific number of hours worked in a day and the binary outcome of whether or not an ocean activity took place.
